---
title: "The direct-effect model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The direct-effect model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_engine <- requireNamespace("netmeta", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4,
  eval = has_engine
)
```

Comparative-effectiveness studies mostly estimate *relative* effects:
drug A versus drug B from an active-comparator design. Each estimate
constrains only the difference between the two drugs' latent direct
effects. The researcher ultimately wants each drug's *direct* effect
versus placebo — and almost none of the evidence speaks to that
directly.

directeffect keeps the two resulting questions separate:

1. **The surface.** Are my comparative estimates internally coherent,
   and what relative structure do they imply? Comparisons alone answer
   this.
2. **The sea level.** Where does that coherent structure sit relative
   to placebo = 0? Only absolute (placebo-anchored) evidence answers
   this.

## Why the separation matters

A null active-comparator estimate does not mean either drug is safe —
only that the two drugs sit at the same height. Three consistent
comparisons and one anchor make the point:

```{r mandatory}
library(directeffect)

comparisons <- data.frame(
  study_id   = c("S1", "S2", "S3"),
  target     = c("A", "A", "B"),
  comparator = c("B", "C", "C"),
  estimate   = c(0.0, 0.4, 0.4),   # A - B is null!
  std_error  = c(0.05, 0.05, 0.05)
)
anchors <- data.frame(
  study_id  = "RCT1",
  drug      = "C",
  reference = "placebo",
  estimate  = 0.3,
  std_error = 0.04
)

de <- direct_effect_network(comparisons, anchors = anchors,
                            effect_measure = "HR")
absolute <- anchor_surface(fit_surface(de, engine = "netmeta"))
absolute$effects[, c("drug", "estimate", "std_error")]
```

The A–B comparison is exactly null, yet both drugs come out at
log HR 0.7 above placebo. The null comparison told us A and B are at
the same height; the anchor on C told us how high that is.

## The workflow on a realistic network

The packaged example data simulate eight fictional drugs with sixteen
head-to-head estimates and two placebo anchors.

```{r construct}
de <- direct_effect_network(
  example_comparisons,
  anchors = example_anchors,
  effect_measure = "HR"
)
de
```

Before fitting anything, inspect the evidence. `check_connectivity()`
reports, per connected component, whether absolute effects are even
identifiable there:

```{r connectivity}
check_connectivity(de)
```

```{r network-plot, eval = has_engine && requireNamespace("ggraph", quietly = TRUE)}
plot_network(de)
```

## Fit the surface, then set the sea level

Surface fitting deliberately ignores the anchors and fixes an arbitrary
reference drug at 0 — only relative positions are identified:

```{r surface}
surface <- fit_surface(de, engine = "netmeta")
surface
```

Anchoring is a distinct, revisable second step. The anchors position
the whole structure against placebo = 0 while keeping their own
uncertainty:

```{r anchor}
absolute <- anchor_surface(surface)
absolute
```

```{r surface-plot, eval = has_engine && requireNamespace("ggplot2", quietly = TRUE)}
plot_effect_surface(absolute)
```

```{r sea-level-plot, eval = has_engine && requireNamespace("ggplot2", quietly = TRUE)}
plot_sea_level(absolute)
```

The same calls draw an unanchored fit too — but then the plot labels
the positions as relative to an arbitrary reference, never as absolute
effects, and `plot_sea_level()` refuses outright: a relative surface
has no sea level.

## Where next

- *netmeta vs. Stan* shows the same surface reconstructed by two
  independent engines — the package's built-in correctness check.
- *Validating the surface* covers edge residuals, cycle consistency,
  and simulation-based recovery.
