---
title: "Getting started with directeffect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with directeffect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_engine <- requireNamespace("netmeta", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4,
  eval = has_engine
)
```

## What directeffect is

Comparative-effectiveness studies mostly estimate *relative* effects:
drug A versus drug B from an active-comparator design. Each such
estimate constrains only the **difference** between the two drugs'
latent direct effects. The researcher ultimately wants each drug's
*direct* effect versus placebo — and almost none of the evidence
speaks to that directly.

directeffect reconstructs the direct effects from the network of
comparative estimates, keeping two questions deliberately separate:

1. **The surface.** Are my comparative estimates internally coherent,
   and what relative structure do they imply? Comparisons alone answer
   this.
2. **The sea level.** Where does that structure sit relative to
   placebo = 0? Only absolute (placebo-anchored) evidence answers this.

The API mirrors the separation: `direct_effect_network()` →
`fit_surface()` → `anchor_surface()`.

## The example data

The package ships a realistic but **entirely simulated** statin
evidence base — real drug names, fake numbers (see `?example_truth`
for the invented truth they were generated from). Twelve head-to-head
estimates among six statins, plus three placebo-controlled anchor
trials, all log hazard ratios:

```{r data}
library(directeffect)

example_comparisons
example_anchors
```

This is the shape real statin evidence has: most trials compare one
active drug against another, while a few landmark placebo trials carry
the absolute information. The input formats are documented in detail
at `?directeffect_formats`.

## 1. Build the network

Construction validates the tables and computes the evidence graph — no
model is fitted yet:

```{r construct}
de <- direct_effect_network(
  example_comparisons,
  anchors = example_anchors,
  effect_measure = "HR"
)
de
```

Before fitting anything, inspect the evidence. `check_connectivity()`
reports, per connected component, whether absolute effects are even
identifiable there — a component with no anchor supports relative
effects only, and the package will never pretend otherwise:

```{r connectivity}
check_connectivity(de)
```

```{r network-plot, eval = has_engine && requireNamespace("ggraph", quietly = TRUE)}
plot_network(de)
```

## 2. Fit the surface

Surface fitting deliberately ignores the anchors and fixes an
arbitrary reference drug at 0 — only relative positions are
identified. Two independent engines are available, `"netmeta"`
(frequentist) and `"stan"` (Bayesian); for networks with one comparison
per study — like this one — they fit the identical model, and they
always return the identical contract. (Multi-arm trials, several
comparisons sharing a `study_id`, are supported by the netmeta engine
only; the Stan engine refuses them rather than mis-treat correlated
rows as independent.)

```{r surface}
surface <- fit_surface(de, engine = "netmeta")
surface
```

## 3. Set the sea level

Anchoring is a distinct, revisable second step. The placebo trials
position the whole structure against placebo = 0 while keeping their
own uncertainty:

```{r anchor}
absolute <- anchor_surface(surface)
absolute
```

`absolute$effects` is a plain data frame (schema at
`?directeffect_formats`) ready for tables or further analysis.

```{r surface-plot, eval = has_engine && requireNamespace("ggplot2", quietly = TRUE)}
plot_effect_surface(absolute)
```

```{r sea-level-plot, eval = has_engine && requireNamespace("ggplot2", quietly = TRUE)}
plot_sea_level(absolute)
```

The same calls draw an unanchored fit too — but then the plot labels
the positions as relative to an arbitrary reference, never as absolute
effects, and `plot_sea_level()` refuses outright: a relative surface
has no sea level.

## 4. Check the evidence

Two quick diagnostics ask whether the comparative estimates actually
agree with one another:

```{r diagnostics}
edge_residuals(surface)

cycle_consistency(surface)
```

*Validating the surface* covers these in depth.

## Where next

- *How directeffect works* explains the method from scratch — no
  background beyond basic statistics assumed — and checks the package
  against the example data's known truth.
- *netmeta vs. Stan* shows the same surface reconstructed by two
  independent engines — the package's built-in correctness check.
- *Validating the surface* covers edge residuals, cycle consistency,
  and simulation-based recovery.
