---
title: "How directeffect works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How directeffect works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_engine <- requireNamespace("netmeta", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4,
  eval = has_engine
)
```

This vignette explains the method from scratch. If you know what an
average and a standard error are, you have all the background you
need.

## The problem: we measure differences, but we want heights

Imagine you want to know how tall three friends are, but your only
measuring stick is broken: it can measure the *difference* between two
people standing back to back, never anyone's actual height.

- Alice vs. Bob: exactly the same height.
- Alice vs. Carol: Alice is 4 cm taller.
- Bob vs. Carol: Bob is 4 cm taller.

From these three measurements you know the *shape* of the lineup
perfectly — Alice and Bob are level, Carol is 4 cm below them — but you
have no idea whether Carol is 150 cm or 190 cm. Every height could be
shifted up or down by the same amount and all three measurements would
still be exactly right.

Drug safety evidence has this structure. Most studies compare one drug
against another drug (drug A vs. drug B), because comparing a drug
against no treatment is often impossible or unethical to run. Each
study measures a **difference between two drugs**, not any drug's own
effect. The number we actually want — the drug's *direct effect*, its
effect versus taking a placebo — is like the actual height.

And this leads to a trap worth naming:

> **A null comparison does not mean two safe drugs.** "Alice vs. Bob:
> same height" is also true when both of them are standing on a table.

## The fix: one measurement against the floor

If just one friend is ever measured against the floor, every height
snaps into place. Say a proper measurement shows Carol is 160 cm. Then
Alice and Bob are both 164 cm — even though no one ever measured
either of them directly.

In drug terms, the "measurement against the floor" is a
placebo-controlled trial. directeffect calls it an **anchor**, and it
calls the floor **sea level** (placebo = 0). The package works in two
deliberate steps:

1. **The surface**: use all the drug-vs-drug comparisons to work out
   the *shape* — every drug's position relative to the others.
2. **The sea level**: use the placebo-controlled anchors to slide that
   whole shape up or down to its true absolute position.

Here is the friends example as data. Effects are on the log hazard
ratio scale — 0 means "no effect", bigger means more harmful — but the
arithmetic is the same as the heights:

```{r toy}
library(directeffect)

comparisons <- data.frame(
  study_id   = c("S1", "S2", "S3"),
  target     = c("A", "A", "B"),
  comparator = c("B", "C", "C"),
  estimate   = c(0.0, 0.4, 0.4),   # A vs B is null!
  std_error  = c(0.05, 0.05, 0.05)
)
anchors <- data.frame(
  study_id  = "RCT1",
  drug      = "C",
  reference = "placebo",
  estimate  = 0.3,               # C measured against the floor
  std_error = 0.04
)

de <- direct_effect_network(comparisons, anchors = anchors,
                            effect_measure = "HR")
absolute <- anchor_surface(fit_surface(de, engine = "netmeta"))
absolute$effects[, c("drug", "estimate", "std_error")]
```

The A-vs-B comparison was exactly null, yet both drugs come out at 0.7
above placebo. The null comparison told us A and B sit at the same
height; the anchor on C told us how high that is.

## What happens when measurements disagree

Real studies are noisy: measure the same pair twice and you get two
slightly different answers, and going around a loop (A vs. B, B vs. C,
C vs. A) may not sum exactly to zero. So the package cannot just chain
measurements together — it has to find the heights that fit *all* the
measurements *at once*, as closely as possible.

The idea is the same as taking an average, with two refinements:

- **Least squares.** Choose the set of heights whose predicted
  differences come closest to all the observed differences. When
  measurements conflict, the answer lands in between, the same way an
  average does.
- **Weighting.** A study with a small standard error is a careful
  measurement; one with a large standard error is a rough one. Each
  measurement counts in proportion to `1 / std_error^2`, so careful
  studies pull harder on the answer.

Anchoring works the same way: each placebo trial proposes "slide the
surface by this much", and the proposals are combined as a weighted
average — the anchors' own uncertainty is kept, so an imprecise placebo
trial gives imprecise absolute effects, as it should.

Every reported estimate carries a standard error and a 95% interval
built from all of this uncertainty together.

## Does it actually work? A test where we know the answer

The package ships a simulated statin evidence base: twelve head-to-head
trials among six statins and three placebo-controlled trials, generated
from an **invented, known truth** (`example_truth`) with realistic
noise added. Real drug names, fake numbers — which is exactly what
makes it useful here: we can run the full method and then grade it
against the answer key.

```{r fit-example}
de <- direct_effect_network(
  example_comparisons,
  anchors = example_anchors,
  effect_measure = "HR"
)
absolute <- anchor_surface(fit_surface(de, engine = "netmeta"))
```

Line the estimates up against the truth the data were generated from:

```{r compare-truth}
truth <- example_truth$theta[match(absolute$effects$drug,
                                   example_truth$drug)]
report <- data.frame(
  drug      = absolute$effects$drug,
  estimated = round(absolute$effects$estimate, 3),
  true      = truth,
  error     = round(absolute$effects$estimate - truth, 3),
  lower     = round(absolute$effects$lower, 3),
  upper     = round(absolute$effects$upper, 3)
)
report[order(report$estimated), ]
```

Three things to check in that table:

```{r grading}
# 1. Every 95% interval contains the true value.
all(report$lower <= report$true & report$true <= report$upper)

# 2. The estimated ordering of the six statins is exactly the true
#    ordering.
cor(report$estimated, report$true, method = "spearman")

# 3. The typical error is about the size the standard errors promised.
sqrt(mean(report$error^2))
```

The estimates miss the truth by a few hundredths on the log scale —
that is the noise we simulated in, not a flaw — every interval covers
the true value, and the package recovers the true ranking of all six
drugs. None of the six was measured against placebo more than once,
and three were never measured against placebo at all; their absolute
effects come entirely from the network plus the anchors.

```{r sea-level-plot, eval = has_engine && requireNamespace("ggplot2", quietly = TRUE)}
plot_sea_level(absolute)
```

This grading is not a one-off: the package's test suite re-runs it (and
many stricter versions of it) on every commit.

## One more safeguard: two independent calculators

Everything above was computed by a frequentist engine (the netmeta
package). directeffect also implements the identical model — identical
likelihood for networks with one comparison per study, like this one —
a second time, in Stan, a completely independent Bayesian engine, and
requires the two to agree on every drug's estimate *and* standard
error, continuously, in its test suite. If two independent
implementations reconstruct the same surface, an error in either would
have to be an error in both. The *netmeta vs. Stan* vignette shows
that comparison on this same example data.
