---
title: "netmeta vs. Stan: two engines, one surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{netmeta vs. Stan: two engines, one surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_engines <- requireNamespace("netmeta", quietly = TRUE) &&
  requireNamespace("rstan", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  eval = has_engines
)
```

The v0.1 goal of directeffect, verbatim from its design: *demonstrate
that the same direct-effect surface is reconstructed independently by
netmeta and Stan.* Both engines fit the identical common-effect model

$$y_k \sim N(\theta_{\text{target}[k]} - \theta_{\text{comparator}[k]},\; se_k^2)$$

— netmeta by frequentist network meta-analysis, Stan by MCMC under a
deliberately weak `normal(0, 5)` prior so the comparison is meaningful.
If two independent implementations agree, an error in either would have
to be an error in both.

## Fit both engines

Both engines sit behind the same `fit_surface()` seam and return the
same fit contract; downstream code cannot tell them apart except via
the `engine` field.

```{r fit}
library(directeffect)

de <- direct_effect_network(
  example_comparisons,
  anchors = example_anchors,
  effect_measure = "HR"
)

fit_nm <- fit_surface(de, engine = "netmeta")
fit_st <- fit_surface(de, engine = "stan", seed = 731, refresh = 0)

fit_nm$effects[, c("drug", "estimate", "std_error")]
```

The Stan fit carries the same core columns plus posterior summaries and
convergence diagnostics:

```{r stan-columns}
fit_st$effects[, c("drug", "estimate", "sd", "rhat", "ess_bulk")]
```

## Compare

`compare_engines()` lines the two fits up per drug:

```{r compare}
comparison <- compare_engines(fit_nm, fit_st)
comparison
```

```{r compare-plot, eval = has_engines && requireNamespace("ggplot2", quietly = TRUE)}
plot_engine_comparison(fit_nm, fit_st)
```

Points on the identity line mean the engines reconstruct the same
surface. The differences here are Monte Carlo noise plus the negligible
shrinkage of the weak prior — both far below anything of scientific
consequence.

## Continuously enforced

This is not just a vignette exercise: the package's test suite
simulates a network under the model's own assumptions and requires
every per-drug difference between the engines to stay below 0.02 on the
log scale, on every commit, in CI. The two implementations validate
each other from day one.

Anchoring works on either engine as well — the Bayesian path refits
with a separate anchored Stan model in which the anchors (not an
arbitrary constraint) identify the absolute location, while the
frequentist path applies a precision-weighted location offset:

```{r anchor}
absolute_nm <- anchor_surface(fit_nm)
absolute_st <- anchor_surface(fit_st, seed = 731, refresh = 0)

data.frame(
  drug = absolute_nm$effects$drug,
  netmeta = round(absolute_nm$effects$estimate, 3),
  stan = round(absolute_st$effects$estimate, 3)
)
```
