---
title: "Validating the surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating the surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_engine <- requireNamespace("netmeta", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4,
  eval = has_engine
)
```

Diagnostics matter more than sophisticated estimation. Before trusting
a fitted surface — let alone an anchored one — ask whether the
comparative estimates are internally coherent, and whether the
estimator would recover a truth you control.

```{r fit}
library(directeffect)

de <- direct_effect_network(
  example_comparisons,
  anchors = example_anchors,
  effect_measure = "HR"
)
fit <- fit_surface(de, engine = "netmeta")
```

## Edge residuals

For every comparison: the observed effect, the effect the fitted
surface predicts (`theta_target - theta_comparator`), and their
difference standardized by the residual's actual standard deviation —
which is smaller than the comparison's standard error, because the
prediction is partly built from the observation itself. The `leverage`
column says how much (at leverage 1, a *bridge* comparison nothing
else can corroborate, the standardized residual is reported as `NA`
rather than as a falsely reassuring 0; this network has no bridges).
Large standardized residuals mark comparisons that conflict with the
rest of the network.

```{r residuals}
residuals <- edge_residuals(fit)
residuals[order(-abs(residuals$standardized_residual)), ][1:5, ]
```

## Cycle consistency

Around any closed loop of comparisons, the observed effects must sum to
zero if they are mutually consistent — walking A→B→C→A has to bring you
back to the same height. `cycle_consistency()` evaluates a *cycle
basis* (every cycle in the network is a combination of basis cycles),
so nothing is enumerated exhaustively; repeat comparisons of a pair are
pooled by precision first.

```{r cycles}
cycles <- cycle_consistency(fit)
cycles
```

```{r cycle-plot, eval = has_engine && requireNamespace("ggplot2", quietly = TRUE)}
plot_cycle_consistency(fit)
```

An inconsistency announces itself. Inject one — flip a single
comparison by 0.5 on the log scale — and both diagnostics flag it:

```{r injected}
broken <- example_comparisons
broken$estimate[3] <- broken$estimate[3] + 0.5
fit_broken <- fit_surface(
  direct_effect_network(broken, effect_measure = "HR"),
  engine = "netmeta"
)

max(abs(edge_residuals(fit_broken)$standardized_residual))
max(abs(cycle_consistency(fit_broken)$z))
```

## Recovery from known truth

The strongest oracle needs no real data at all: simulate a network
where the truth is known, fit it, and measure recovery.
`simulate_direct_effect_network()` retains the generating truth, and
`validate_recovery()` reports bias, RMSE, interval coverage, and rank
correlation from the fit contract alone — it works identically for
either engine.

```{r recovery}
simulation <- simulate_direct_effect_network(
  n_drugs = 20, n_comparisons = 80, n_anchors = 3,
  heterogeneity = 0, seed = 42, se_range = c(0.03, 0.08)
)
recovered <- fit_surface(simulation$network, engine = "netmeta")
validate_recovery(recovered, simulation)
```

In the low-noise regime the bias is near zero and coverage is near
nominal; the package's test suite pins both down permanently. Together
the three oracles — hand-computable examples, cross-engine agreement
(see *netmeta vs. Stan*), and simulation truth — are what let the
surface be trusted.
